---
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
repo_root <- normalizePath(".", winslash = "/", mustWork = TRUE)

require_package <- function(package) {
  if (!requireNamespace(package, quietly = TRUE)) {
    stop("Install the R package '", package, "' to render README.md", call. = FALSE)
  }
}

require_package("jsonlite")

pkg <- jsonlite::fromJSON(file.path(repo_root, "package.json"), simplifyVector = FALSE)

strip_dot_slash <- function(x) sub("^\\./", "", x)

expand_pi_entries <- function(entries, predicate) {
  entries <- unlist(entries, use.names = FALSE)
  entries <- strip_dot_slash(entries)
  includes <- entries[!startsWith(entries, "!")]
  excludes <- strip_dot_slash(sub("^!", "", entries[startsWith(entries, "!")]))

  expand_one <- function(entry) {
    if (grepl("[*?\\[]", entry)) {
      return(Sys.glob(entry))
    }
    if (dir.exists(entry)) {
      children <- list.files(entry, full.names = TRUE, recursive = FALSE, include.dirs = TRUE)
      matches <- c(entry, children)
      return(matches[file.info(matches)$isdir %in% TRUE])
    }
    entry
  }

  paths <- unique(unlist(lapply(includes, expand_one), use.names = FALSE))
  excluded <- unique(unlist(lapply(excludes, expand_one), use.names = FALSE))
  paths <- setdiff(paths, excluded)
  paths <- paths[file.info(paths)$isdir %in% TRUE]
  sort(paths[vapply(paths, predicate, logical(1))])
}

is_skill_dir <- function(path) file.exists(file.path(path, "SKILL.md"))
is_extension_dir <- function(path) {
  file.exists(file.path(path, "index.ts")) || file.exists(file.path(path, "index.js"))
}

frontmatter_value <- function(lines, key) {
  if (!length(lines) || lines[[1]] != "---") return(NA_character_)
  ends <- which(lines[-1] == "---")
  if (!length(ends)) return(NA_character_)
  block <- lines[seq.int(2, ends[[1]])]
  hit <- grep(paste0("^", key, ":[[:space:]]*"), block, value = TRUE)
  if (!length(hit)) return(NA_character_)
  value <- sub(paste0("^", key, ":[[:space:]]*"), "", hit[[1]])
  sub("^[\"']?(.*?)[\"']?$", "\\1", value, perl = TRUE)
}

read_skill_meta <- function(path) {
  lines <- readLines(file.path(path, "SKILL.md"), warn = FALSE, encoding = "UTF-8")
  name <- frontmatter_value(lines, "name")
  description <- frontmatter_value(lines, "description")
  if (is.na(name) || !nzchar(name)) name <- basename(path)
  if (is.na(description)) description <- ""
  list(name = name, description = description)
}

extension_descriptions <- c(
  goals = "Codex-style `/goals` and `/goal` session goal loop extension.",
  rlm = "Single-controller long-context tool with externalized inspection, bounded opt-in recursion, system `Rscript` evaluation, and DuckDB-backed parquet sampling."
)

extension_paths <- expand_pi_entries(pkg$pi$extensions, is_extension_dir)
skill_paths <- expand_pi_entries(pkg$pi$skills, is_skill_dir)
skills <- lapply(skill_paths, read_skill_meta)
skills <- skills[order(vapply(skills, `[[`, character(1), "name"))]

cat_bullets <- function(rows) {
  invisible(lapply(rows, function(row) {
    if (nzchar(row$description)) {
      cat(sprintf("- `%s` — %s\n", row$name, row$description))
    } else {
      cat(sprintf("- `%s`\n", row$name))
    }
  }))
}
```

# pi-skills

<!-- README.md is generated from README.Rmd. Edit README.Rmd and run `npm run render:readme`. -->

Personal, skills-first Pi package for `sounkou-bioinfo`.

It includes domain skills, the user-wide `/goals` extension, and a lean RLM tool that uses system R directly.

## Install

```sh
pi install git:github.com/sounkou-bioinfo/pi-skills
```

After installing or updating in an interactive Pi session:

```text
/reload
```

## Active contents

The extension and skill lists below are generated from `package.json` plus skill frontmatter.

### Extensions

```{r extensions, results='asis', echo=FALSE}
extension_rows <- lapply(extension_paths, function(path) {
  name <- basename(path)
  description <- if (name %in% names(extension_descriptions)) unname(extension_descriptions[[name]]) else ""
  list(name = name, description = description)
})
cat_bullets(extension_rows)
```

### Orchestration discipline

This package does not ship a subagent/scout/plan/recon swarm and does not create worktrees. The RLM keeps long context outside the model and uses one controller to inspect it. Its default `auto` mode does not recurse; `mode=decompose` is an explicit, shallow, budgeted exception for materially independent contradictions.

Only one RLM run and one child model process are active at a time; at most four active/queued runs are retained. Additional background runs and recursive children queue. This serialization is a hard safety boundary, not a tuning default.

Inline context is capped at 5M characters. Eager text files are capped at 20MB and eager CSV/JSON files at 10MB. Directory contexts keep a Git-aware manifest of up to 20,000 files, eagerly load at most 5MB of prioritized text, and expose bounded lazy reads for omitted text files.

### Skills

```{r skills, results='asis', echo=FALSE}
cat_bullets(skills)
```

## RLM and R runtime direction

`r_eval` now targets real system R, not webR. Configure it with:

- tool param `rBin`, or env `PI_RLM_R_BIN` (default: `Rscript`)
- tool param `rLibPaths`, or env `PI_RLM_R_LIBS` (`:`/`;` separated)
- tool param `rRepos`, or env `PI_RLM_R_REPOS` (default: `https://cloud.r-project.org`)

Inside `r_eval`, use `install_r_packages()`, `context_load()`, `context_r_load_code()`, `save_plot()`, `FINAL()`, and `FINAL_VAR()`. `rlm_call()` is available only when recursion was explicitly enabled with `mode=decompose`.

Parquet context is sampled through `@duckdb/node-api`; full analysis remains available through `context_load()` in system R. This lines up with the DuckDB-native direction for DuckHTS, `ducknng`, and `ducktinycc` without eagerly retaining an entire Parquet table in Node.

The longer-term worker design is tracked in `docs/rlm-r-nng-arrow-design.md`: persistent system-R workers over NNG/nanoarrow, with DuckDB extensions as first-class runtime peers.

## Development

```sh
npm install
npm run typecheck
npm run test:rlm
npm run render:readme
```

Runtime Pi APIs are peer dependencies supplied by Pi itself.

## License

MIT. See `LICENSE`.
