---
output:
  github_document:
    html_preview: false
---

```{r setup, include=FALSE}
repo_root <- normalizePath(".", winslash = "/", mustWork = TRUE)

require_package <- function(package) {
  if (!requireNamespace(package, quietly = TRUE)) {
    stop("Install the R package '", package, "' to render README.md", call. = FALSE)
  }
}

require_package("jsonlite")

pkg <- jsonlite::fromJSON(file.path(repo_root, "package.json"), simplifyVector = FALSE)

strip_dot_slash <- function(x) sub("^\\./", "", x)

expand_pi_entries <- function(entries, predicate) {
  entries <- unlist(entries, use.names = FALSE)
  entries <- strip_dot_slash(entries)
  includes <- entries[!startsWith(entries, "!")]
  excludes <- strip_dot_slash(sub("^!", "", entries[startsWith(entries, "!")]))

  expand_one <- function(entry) {
    if (grepl("[*?\\[]", entry)) {
      return(Sys.glob(entry))
    }
    if (dir.exists(entry)) {
      children <- list.files(entry, full.names = TRUE, recursive = FALSE, include.dirs = TRUE)
      matches <- c(entry, children)
      return(matches[file.info(matches)$isdir %in% TRUE])
    }
    entry
  }

  paths <- unique(unlist(lapply(includes, expand_one), use.names = FALSE))
  excluded <- unique(unlist(lapply(excludes, expand_one), use.names = FALSE))
  paths <- setdiff(paths, excluded)
  paths <- paths[file.exists(paths) | dir.exists(paths)]
  sort(paths[vapply(paths, predicate, logical(1))])
}

is_skill_dir <- function(path) file.exists(file.path(path, "SKILL.md"))
is_extension_path <- function(path) {
  if (dir.exists(path)) {
    return(file.exists(file.path(path, "index.ts")) || file.exists(file.path(path, "index.js")))
  }
  grepl("[.](ts|js)$", path)
}

frontmatter_value <- function(lines, key) {
  if (!length(lines) || lines[[1]] != "---") return(NA_character_)
  ends <- which(lines[-1] == "---")
  if (!length(ends)) return(NA_character_)
  block <- lines[seq.int(2, ends[[1]])]
  hit <- grep(paste0("^", key, ":[[:space:]]*"), block, value = TRUE)
  if (!length(hit)) return(NA_character_)
  value <- sub(paste0("^", key, ":[[:space:]]*"), "", hit[[1]])
  sub("^[\"']?(.*?)[\"']?$", "\\1", value, perl = TRUE)
}

read_skill_meta <- function(path) {
  lines <- readLines(file.path(path, "SKILL.md"), warn = FALSE, encoding = "UTF-8")
  name <- frontmatter_value(lines, "name")
  description <- frontmatter_value(lines, "description")
  if (is.na(name) || !nzchar(name)) name <- basename(path)
  if (is.na(description)) description <- ""
  list(name = name, description = description)
}

extension_descriptions <- c(
  `anti-slop` = "Tree-sitter-only, configurable R/C audit for redundant code, private-helper call counts and justification, ambiguous length conditions, empty branches, and host-unsafe C runtime assertions.",
  `background-tasks` = "Named detachable shell tasks with bounded external logs, status/kill tools, a focused TUI dock, and one completion wakeup.",
  `context-budget` = "Transiently caps large built-in inspection results in model context while preserving full stored/UI results and deterministic recovery guidance.",
  goals = "Session goal loop with stable policy, transient goal context, compact continuations, and evidence-gated completion.",
  memory = "Append-only Semantic-SQL memory with cache-safe task retrieval, bounded summary frontiers, historical `as_of`, graph traversal, and DuckDB FTS.",
  rlm = "Detached-by-default single-controller long-context runs with completion wakeups, bounded opt-in recursion, system `Rscript` evaluation, and DuckDB-backed parquet sampling."
)

extension_paths <- expand_pi_entries(pkg$pi$extensions, is_extension_path)
skill_paths <- expand_pi_entries(pkg$pi$skills, is_skill_dir)
skills <- lapply(skill_paths, read_skill_meta)
skills <- skills[order(vapply(skills, `[[`, character(1), "name"))]

cat_bullets <- function(rows) {
  invisible(lapply(rows, function(row) {
    if (nzchar(row$description)) {
      cat(sprintf("- `%s` — %s\n", row$name, row$description))
    } else {
      cat(sprintf("- `%s`\n", row$name))
    }
  }))
}
```

# pi-skills

<!-- README.md is generated from README.Rmd. Edit README.Rmd and run `npm run render:readme`. -->

Personal, skills-first Pi package for `sounkou-bioinfo`.

It includes domain skills, permanent semantic memory, detachable background tasks, the user-wide `/goals` extension, and a lean RLM tool that uses system R directly.

## Install

```sh
pi install git:github.com/sounkou-bioinfo/pi-skills
```

After installing or updating in an interactive Pi session:

```text
/reload
```

## Active contents

The extension and skill lists below are generated from `package.json` plus skill frontmatter.

### Extensions

```{r extensions, results='asis', echo=FALSE}
extension_rows <- lapply(extension_paths, function(path) {
  name <- sub("[.](ts|js)$", "", basename(path))
  description <- if (name %in% names(extension_descriptions)) unname(extension_descriptions[[name]]) else ""
  list(name = name, description = description)
})
cat_bullets(extension_rows)
```

### Anti-slop AST audit

`anti_slop` and `/anti-slop path/to/source.R` audit one R/C source file or a directory with the vendored Tree-sitter analyzer. A Git directory scans its tracked R/C files beneath that path; a non-Git directory recurses over recognized source suffixes. It counts direct calls to each top-level R private helper across that analysis scope and asks for its distinct invariant/effect, flags condition sprawl and ambiguous `length()` truthiness, and inspects narrow redundant/host-unsafe C patterns. The extension deliberately does not use regex or a substitute parser: unavailable R grammar packages and parse errors are explicit results. Runtime requirements are `treesitter`, `treesitter.r` for R, and [treesitter.c](https://github.com/sounkou-bioinfo/treesitter.c) for C; optional JSON configuration and machine-readable output use `jsonlite`. See the `r-c-anti-slop` skill for its narrow rule scope and review workflow.

### Orchestration discipline

This package does not ship a subagent/scout/plan/recon swarm and does not create worktrees. The RLM keeps long context outside the model and uses one controller to inspect it. Its default `auto` mode does not recurse; `mode=decompose` is an explicit, shallow, budgeted exception for materially independent contradictions.

Skill discovery uses progressive disclosure, but names and descriptions are always in the system prompt. The inventory therefore keeps one distinct trigger per skill and delegates shared mechanics to a single authority instead of multiplying near-duplicate skills.

The `context-budget` hook prevents broad `read`, `bash`, `grep`, `find`, or `ls` results from consuming the context window immediately after compaction. Each result is capped at 12 KiB with deterministic head/tail evidence; an aggregate 64 KiB budget retains newest inspection results and replaces older bodies with rerunnable receipts. Complete tool results remain stored and visible. Configure `PI_CONTEXT_TOOL_RESULT_BYTES` (4096–51200) and `PI_CONTEXT_TOOL_RESULTS_TOTAL_BYTES` (16384–524288).

Only one RLM run and one child model process are active at a time; at most four active/queued runs are retained. Additional background runs and recursive children queue. RLM starts detached by default, keeps the Pi session interactive, and emits one bounded completion wakeup; set `async=false` only for a short blocking call. This serialization is a hard safety boundary, not a tuning default.

Long-running shell commands use `bg_run`: output stays in bounded external task files while Pi remains interactive, and completion emits one status/path wakeup. This avoids holding a foreground tool call open or repeatedly injecting CI progress into model context.

### Permanent memory

The `memory` extension keeps one append-only SQLite authority at `~/.pi/agent/memory.sqlite` (override with `PI_MEMORY_DB`). SQLite runs in WAL mode; each Pi process owns an in-memory DuckDB connection attached through DuckDB's SQLite storage extension. Transactions and RDF-shaped statements are the small physical core. Notes, current facts, history, summaries, graph edges, pending compression, and historical `as_of` projections are SQL views or recursive SQL queries.

The system prompt contains only stable memory policy. Once per user turn, the extension builds a small task-conditioned projection from a four-row summary frontier plus at most four exact FTS matches. The `context` hook inserts that projection immediately before the current user message, after the prior transcript. It is not persisted, does not accumulate, and remains frozen through that turn's tool loop. A memory write therefore changes retrieval on the next user turn without mutating the early system-prompt cache prefix.

Repeated `(graph, subject, predicate)` statements create inspectable versions rather than rewriting history. `memory sql` selects an explicit transaction or timestamp and exposes `as_of_statement`, `as_of_note`, `as_of_summary`, `node_to_node_statement`, `node_to_value_statement`, `memory_block`, `pending_summary`, and `wake`. Its dedicated read-only connection disables local and network filesystem access, locks DuckDB configuration, and bounds query text, memory, threads, runtime, and returned rows. `recall` maintains a process-local DuckDB FTS projection over exact notes; SQLite remains authoritative.

Summary nodes and their `memory:left` / `memory:right` statements form a graph that the agent can walk with recursive SQL or `zoom`. The `graph` column leaves named-graph/fork projections schema-compatible, but this package does not invent branch-head or merge policy before a current consumer needs it.

The few-base-tables/many-semantic-views design follows [INCATools Semantic-SQL](https://github.com/INCATools/semantic-sql). The append-only log, bounded fading-detail context, and navigable summary hierarchy are independently implemented concepts inspired by [VictorTaelin/OptMem](https://github.com/VictorTaelin/OptMem); no OptMem source is included.

Inline context is capped at 5M characters. Eager text files are capped at 20MB and eager CSV/JSON files at 10MB. Directory contexts keep a Git-aware manifest of up to 20,000 files, eagerly load at most 5MB of prioritized text, and expose bounded lazy reads for omitted text files.

### Skills

```{r skills, results='asis', echo=FALSE}
cat_bullets(skills)
```

## RLM and R runtime direction

`r_eval` now targets real system R, not webR. Configure it with:

- tool param `rBin`, or env `PI_RLM_R_BIN` (default: `Rscript`)
- tool param `rLibPaths`, or env `PI_RLM_R_LIBS` (`:`/`;` separated)
- tool param `rRepos`, or env `PI_RLM_R_REPOS` (default: `https://cloud.r-project.org`)

Inside `r_eval`, use `install_r_packages()`, `context_load()`, `context_r_load_code()`, `save_plot()`, `FINAL()`, and `FINAL_VAR()`. `rlm_call()` is available only when recursion was explicitly enabled with `mode=decompose`.

Parquet context is sampled through `@duckdb/node-api`; full analysis remains available through `context_load()` in system R. This lines up with the DuckDB-native direction for DuckHTS, `ducknng`, and `ducktinycc` without eagerly retaining an entire Parquet table in Node.

The longer-term worker design is tracked in `docs/rlm-r-nng-arrow-design.md`: persistent system-R workers over NNG/nanoarrow, with DuckDB extensions as first-class runtime peers.

## Development

```sh
npm install
npm run typecheck
npm run test
npm run test:memory
npm run test:rlm
npm run render:readme
npm pack --dry-run
```

Runtime Pi APIs are peer dependencies supplied by Pi itself.

## License

MIT. See `LICENSE`.
