---
title: "Memory, Lode, and LeanContext"
subtitle: "What pi-skills has today, what the alternatives are, and what integrating them would mean"
date: "2026-08-16"
output:
  litedown::html_format:
    meta:
      css: ["@default", "@article", "memory-lode-design.css"]
    options:
      toc: true
      embed_resources: ["https", "local"]
---

```{r setup, include=FALSE}
litedown::reactor(
  echo = FALSE, message = FALSE, warning = FALSE,
  dev = Sys.getenv("FIGDEV", "svg"), dpi = 72,
  fig.width = 13.6, fig.height = 5
)

## Validated light-mode palette (dataviz reference instance)
SURF <- "#fcfcfb"; INK <- "#0b0b0b"; INK2 <- "#52514e"; MUT <- "#898781"
GRID <- "#e1e0d9"; BASE <- "#c3c2b7"
BLU  <- "#2a78d6"; BLU100 <- "#cde2fb"; BLU200 <- "#9ec5f4"; BLU550 <- "#1c5cab"
ORA  <- "#eb6834"

## Figures are sized in pixels, not inches: the svg device writes
## inches * 72 px, so 13.6in == 979px, matching the CSS content width.
## ps = 15 keeps label text at 13-15px on screen instead of 8px.
blank_plot <- function(xlim, ylim, mar = c(0.2, 0.2, 0.2, 0.2)) {
  par(mar = mar, bg = SURF, family = "sans", xpd = NA, ps = 15)
  plot.new(); plot.window(xlim = xlim, ylim = ylim)
}

## Read a numeric constant straight out of the TypeScript sources, so the
## figures cannot drift away from the implementation.
const_num <- function(path, name) {
  src <- paste(readLines(file.path("..", path), warn = FALSE), collapse = "\n")
  m <- regmatches(src, regexpr(paste0(name, "\\s*=\\s*[0-9_]+"), src))
  if (!length(m)) return(NA_integer_)
  as.integer(gsub("[^0-9]", "", m))
}

NOTE_BYTES  <- const_num("extensions/memory/store.ts", "MAX_MEMORY_BYTES")
WAKE_ROWS   <- const_num("extensions/memory/index.ts", "HISTORY_LINE_BUDGET")
MATCH_ROWS  <- const_num("extensions/memory/index.ts", "TASK_MATCH_LIMIT")

## Measured size of the stable system-prompt policy the two extensions append.
policy_bytes <- function(path, marker) {
  x <- readLines(file.path("..", path), warn = FALSE)
  i <- grep(marker, x, fixed = TRUE)
  if (!length(i)) return(0L)
  j <- grep("^`;$|`;\\s*$", x[i:length(x)])[1] + i - 1L
  sum(nchar(x[i:j], type = "bytes"))
}
MEM_POLICY  <- policy_bytes("extensions/memory/index.ts", "PERMANENT MEMORY POLICY")
GOAL_POLICY <- policy_bytes("extensions/goals/index.ts", "PI GOAL POLICY")
```

## 0. Summary

Five systems are on the table. Four are external; one is ours.

| System | Substrate | Written by | History | Verdict for us |
|---|---|---|---|---|
| **pi-skills `memory`** | SQLite/WAL via DuckDB, append-only statements | agent | per-field, full `as_of` | **the authority — extend it** |
| **Lode Coding** | `lode/` markdown tree | agent, on request | git only, per file | take the *vocabulary*, reject the files |
| **LeanContext** | `CONSTITUTION.md` + `MANIFEST.md` + `.context/` | human | none (manual `Status:` line) | take the *tier semantics* and the admission test |
| **Claude Code CLAUDE.md / rules** | markdown hierarchy | human | git only | take `paths:` glob triggering |
| **Claude Code auto memory** | `~/.claude/projects/<repo>/memory/` | Claude | `modified` timestamp | take the enforced index budget |
| **Anthropic CMA memory stores** | server-side documents | either | immutable per-document versions + **redact** | take **redaction** |

Decisions already settled in discussion:

- No `lode/` directory, no markdown sprawl, no VFS.
- Lode becomes a **vocabulary and a graph scope inside `memory`**, not a second store.
- Project knowledge reaches git as a **generated single-file export**, never as the authority.

---

## 1. What we have today: the `memory` extension

### 1.1 Physical core

Three tables. Everything else is a view.

```
memo.transactions (transaction_id, transaction_time, operation, session_id, receipt)
memo.statements   (transaction_id, ordinal, stanza, subject, predicate,
                   object, value, datatype, language, graph)
memo.memory_meta  (schema_version = "1")   memo.memory_counter (transaction)
```

`extensions/memory/schema.sql` · `extensions/memory/store.ts`

### 1.2 The one semantic idea: slots supersede, they do not overwrite

A **slot** is `(graph, subject, predicate)`. Appending a statement to an occupied
slot supersedes the old value without rewriting it, so every field carries its own
validity interval and the whole store can be replayed at any transaction.

```{r fig-asof, fig.height=5.2}
slots <- list(
  list(lab = "lode:covers",      tx = c(3, 11, 18)),
  list(lab = "lode:verifiedRev", tx = c(5, 14)),
  list(lab = "lode:status",      tx = c(7))
)
AS_OF <- 12; XMAX <- 21

blank_plot(c(-6, XMAX + 0.6), c(0.2, length(slots) + 1.15), mar = c(2.6, 0.2, 0.2, 0.2))

## transaction axis
segments(0, 0.55, XMAX, 0.55, col = BASE, lwd = 1)
at <- c(0, 5, 10, 15, 20)
segments(at, 0.55, at, 0.42, col = BASE, lwd = 1)
text(at, 0.06, at, col = MUT, cex = 0.86)
text(XMAX / 2, -0.28, "transaction_id", col = MUT, cex = 0.88)

for (i in seq_along(slots)) {
  y <- length(slots) - i + 1.25
  s <- slots[[i]]
  ends <- c(s$tx[-1], XMAX)
  text(-0.55, y, s$lab, adj = 1, col = INK2, cex = 0.98, family = "mono")
  for (k in seq_along(s$tx)) {
    live <- s$tx[k] <= AS_OF && ends[k] > AS_OF
    segments(s$tx[k], y, ends[k], y,
             col = if (live) BLU else BASE, lwd = if (live) 4.5 else 2,
             lend = "butt")
    points(s$tx[k], y, pch = 21, bg = if (live) BLU else SURF,
           col = if (live) BLU else MUT, cex = 1.15, lwd = 1.4)
  }
}

segments(AS_OF, 0.72, AS_OF, length(slots) + 0.82, col = ORA, lwd = 1.6, lty = 2)
text(AS_OF + 0.25, length(slots) + 0.92, "as_of = 12", col = ORA, cex = 0.95, adj = 0)

legend(-6, length(slots) + 1.15, bty = "n", horiz = TRUE, cex = 0.88,
       text.col = INK2, seg.len = 1.6, lwd = c(4.5, 2), col = c(BLU, BASE),
       legend = c("current at as_of", "superseded (retained)"))
```

One `UPDATE` of a single temp row retargets every `as_of_*` view at once:

| View | What it gives | Where |
|---|---|---|
| `statement_history` | `valid_from` / `valid_to` per slot, via `lead()` | `schema.sql:71` |
| `current_statement` | the head of every slot | `schema.sql:87` |
| `as_of_statement` | the head **as of** a chosen transaction | `session.sql:9` |
| `memory_query_context` | one temp row (`as_of_transaction`, `line_budget`) | `session.sql:3` |

`session.sql` runs on both the write connection (`store.ts:129`) and the isolated
read-only `sql` connection (`store.ts:313`), so `memory sql` sees the same views.

### 1.3 Notes, summaries, and the wake frontier

| Layer | Mechanism | Bound |
|---|---|---|
| Note | stanza `memory:note/<tx>`, one line | **`{r} NOTE_BYTES` UTF-8 bytes** |
| Summary tree | every aligned power-of-two block of the note stream, compressed by `nap` | source-hash checked |
| Wake frontier | recursive binary walk biased toward the present | `line_budget` rows |
| Recall | process-local DuckDB FTS over `note_statement`, bm25 | rebuilt per watermark |
| `sql` | read-only connection, no filesystem or network, bounded rows/memory/threads/runtime | hard |

The frontier below is not a sketch — it runs the same recursion as the `wake_frontier`
CTE in `session.sql`, for `{r} WAKE_ROWS` slots (`HISTORY_LINE_BUDGET`) over 23 notes.
Old ranges collapse into one summary line; detail sharpens toward the present.

```{r fig-wake, fig.height=3.6}
wake_frontier <- function(note_count, line_budget) {
  top <- 1; while (top < note_count) top <- top * 2
  out <- list()
  walk <- function(a, b, slots) {
    if (a >= note_count) return(invisible(NULL))
    emit <- (b - a == 1) || (b <= note_count && slots <= 1)
    if (emit) { out[[length(out) + 1L]] <<- c(a, b); return(invisible(NULL)) }
    if ((b - a) > 1 && (b > note_count || slots > 1)) {
      mid <- (a + b) %/% 2
      lslots <- if (note_count <= mid) slots else max(1, slots %/% 3)
      rslots <- if (slots > 1) max(1, slots - max(1, slots %/% 3)) else 1
      walk(a, mid, lslots); walk(mid, b, rslots)
    }
  }
  walk(0, top, line_budget)
  do.call(rbind, out)
}

N <- 23
fr <- wake_frontier(N, WAKE_ROWS)

blank_plot(c(-1.4, N + 2.6), c(-1.25, 1.55), mar = c(0.2, 0.2, 0.2, 0.2))
for (i in seq_len(nrow(fr))) {
  a <- fr[i, 1]; b <- min(fr[i, 2], N); single <- (b - a) == 1
  rect(a + 0.08, -0.35, b - 0.08, 0.55,
       col = if (single) BLU else BLU100,
       border = if (single) BLU550 else BLU, lwd = 1.2)
  lab <- if (single) paste0("#", a) else paste0("#", a, "–", b - 1)
  text((a + b) / 2, 0.10, lab, cex = 0.86,
       col = if (single) "#ffffff" else BLU550)
}
segments(0, -0.62, N, -0.62, col = BASE)
text(c(0, N), -0.92, c("oldest note", "newest note"), col = MUT, cex = 0.86,
     adj = c(0, 1))
text(N / 2, -0.92, paste0(nrow(fr), " rows for ", N, " notes"),
     col = MUT, cex = 0.86)
legend(-1.4, 1.55, bty = "n", horiz = TRUE, cex = 0.88, text.col = INK2,
       pch = 22, pt.cex = 1.5, pt.lwd = 1.2,
       pt.bg = c(BLU100, BLU), col = c(BLU, BLU550),
       legend = c("summary of a collapsed range", "raw note, verbatim"))
```

### 1.4 Where context actually enters the model

This is the part most external systems get wrong, and ours already gets right:
the cached prefix never moves, and everything task-conditioned is transient.

```{r fig-context, fig.height=6}
bands <- data.frame(
  lab  = c("system prompt — stable policy",
           "transcript (grows)",
           "<memory_projection>",
           "<active_goal>",
           "current user message"),
  note = c("memory + goal policy, measured below",
           "prior turns, untouched",
           paste0(WAKE_ROWS, " frontier rows + ≤", MATCH_ROWS, " FTS matches"),
           "objective + turn counters",
           "the task"),
  kind = c("stable", "stable", "transient", "transient", "now"),
  stringsAsFactors = FALSE
)
h <- 1; gap <- 0.34
n <- nrow(bands)
tops <- rev(seq_len(n)) * (h + gap)

blank_plot(c(0, 100), c(0.15, max(tops) + h + 0.7), mar = c(0.2, 0.2, 0.2, 0.2))
fill <- c(stable = BLU100, transient = "#ffffff", now = "#f1f0ea")
edge <- c(stable = BLU,    transient = ORA,       now = BASE)

for (i in seq_len(n)) {
  y <- tops[i]; k <- bands$kind[i]
  rect(12, y, 62, y + h, col = fill[[k]], border = edge[[k]], lwd = 1.4,
       lty = if (k == "transient") 2 else 1)
  text(14, y + h / 2, bands$lab[i], adj = 0, col = INK, cex = 0.96,
       family = if (startsWith(bands$lab[i], "<")) "mono" else "sans")
  segments(62, y + h / 2, 65, y + h / 2, col = GRID)
  text(66, y + h / 2, bands$note[i], adj = 0, col = MUT, cex = 0.88)
}

## cached-prefix bracket
yb <- tops[2]; yt <- tops[1] + h
segments(9.5, yb, 9.5, yt, col = BLU, lwd = 1.4)
segments(9.5, c(yb, yt), 11, c(yb, yt), col = BLU, lwd = 1.4)
text(8.6, (yb + yt) / 2, "cached prefix", srt = 90, col = BLU550, cex = 0.88)

## transient bracket
yb2 <- tops[4]; yt2 <- tops[3] + h
segments(9.5, yb2, 9.5, yt2, col = ORA, lwd = 1.4)
segments(9.5, c(yb2, yt2), 11, c(yb2, yt2), col = ORA, lwd = 1.4)
text(8.6, (yb2 + yt2) / 2, "rebuilt per turn", srt = 90, col = ORA, cex = 0.88)

text(12, max(tops) + h + 0.45, "request order ↓", adj = 0, col = MUT, cex = 0.88)
```

The projection is keyed on `sessionId : userTimestamp : sha256(prompt)`
(`index.ts:212`), frozen for the whole tool loop, never persisted, and replaced
rather than accumulated. A write therefore changes retrieval on the **next** user
turn without touching the cache prefix. Around it, `context-budget` caps each
built-in inspection result at 12 KiB and the aggregate at 64 KiB, freezing each
result's fate the first time it is seen so prompt caching survives.

### 1.5 Limits worth remembering

| Field | Cap |
|---|---|
| note text | `{r} NOTE_BYTES` bytes, one line |
| stanza / subject | 300 bytes |
| predicate / graph | 200 bytes |
| object | 500 bytes |
| statement `value` | one line (no byte cap enforced) |
| `memory:system` graph | reserved for derived summaries |
| `SCHEMA_VERSION` | `"1"`, strict equality — a bump breaks existing DBs |

Adding **new views with new names** needs no version bump. Lode needs no new tables.

---

## 2. What Claude Code actually runs (this session, right now)

Four separate systems ship under the name "memory". They are not variants of each other.

### 2.1 CLAUDE.md hierarchy — human instructions

| Scope | Location | Shared with |
|---|---|---|
| Managed policy | `/etc/claude-code/CLAUDE.md` (Linux) | everyone on the machine, cannot be excluded |
| User | `~/.claude/CLAUDE.md` | you, all projects |
| Project | `./CLAUDE.md` or `./.claude/CLAUDE.md` | the team, via git |
| Local | `./CLAUDE.local.md` | you, this project (gitignored) |

- Files are **concatenated**, root-of-tree downward, so the most specific is read last. Subdirectory files load lazily when Claude reads a file there.
- `@path/to/file` imports, relative to the importing file, **max 4 hops**; backticked paths are not imported. Imports do not save context — they still load at launch.
- `.claude/rules/*.md` with `paths:` glob frontmatter load **only when Claude reads a matching file**. This is the one genuinely lazy tier.
- Target **under 200 lines**. `claudeMdExcludes` skips other teams' files in a monorepo.
- Delivered **as a user message after the system prompt**, explicitly not as system prompt. Not enforcement: *"to block an action regardless of what Claude decides, use a PreToolUse hook."*

### 2.2 Auto memory — the system I am running under

Storage for this very session: `/root/.claude/projects/-root-pi-skills/memory/` —
currently **empty**; nothing has been saved for this repo yet.

| Property | Value |
|---|---|
| Scope | derived from the **git repository** — all worktrees and subdirectories share one directory |
| Sync | machine-local; never shared across machines or cloud |
| Loaded each session | `MEMORY.md` only, **first 200 lines or 25 KB**, whichever comes first |
| Topic files | read on demand with ordinary file tools |
| Retention | excluded from the transcript cleanup sweep |
| Subagents | separate directory; a fork inherits the parent's |
| Over budget | the write succeeds, then the harness **returns an error telling Claude to rewrite the index** |
| Freshness | a `modified` ISO-8601 timestamp is stamped into frontmatter on write |

The instructions I carry in my system prompt right now, verbatim in shape:

```markdown
---
name: <short-kebab-case-slug>
description: <one-line summary, used to decide relevance during recall>
metadata:
  type: user | feedback | project | reference
---

<the fact; for feedback/project, follow with **Why:** and **How to apply:** lines.
 Link related memories with [[their-name]].>
```

with these rules attached:

- **One file, one fact.** Types: `user` (who you are), `feedback` (how I should work, with the why), `project` (ongoing work not derivable from code or git), `reference` (URLs, dashboards, tickets).
- Link liberally with `[[name]]`; a link to a memory that does not exist yet is a marker, not an error.
- After writing, add **one line** to `MEMORY.md`: `- [Title](file.md) — hook`. The index is loaded every session; never put content there.
- Check for an existing file covering the fact and update it rather than duplicating. Delete memories that turn out wrong.
- **Do not save what the repo already records** — code structure, past fixes, git history, CLAUDE.md. If asked to save one of those, ask what was non-obvious and save that instead.
- Recalled memories arrive inside `<system-reminder>` blocks: background context, not instructions, and true only as of when they were written. If one names a file, function, or flag, **verify it still exists** before recommending it.
- Convert relative dates to absolute.

Read that list against ours and the overlap is striking: a knowledge-boundary rule,
one-fact-per-record, a bounded always-loaded index, a staleness warning, and a
dedupe-before-write rule — the same five things LeanContext specifies in prose and
we want to specify in SQL.

### 2.3 The API memory tool (`memory_20250818`)

Client-side. Claude issues `view` / `create` / `str_replace` / `insert` / `delete` /
`rename` against a `/memories` directory; **you implement the backend**. Pairs with
context editing (`clear_tool_uses_20250919`) and server-side compaction. No history.

This is the only reason a VFS over our SQLite store would ever exist — it is the
file-shaped interface an API-side agent expects. We are not targeting that surface,
so it stays unbuilt.

### 2.4 Managed-Agents memory stores

The only Anthropic memory surface with real history.

| Object | ID | Notes |
|---|---|---|
| Store | `memstore_` | workspace-scoped, FUSE-mounted at `/mnt/memory/<name>/`, `read_only` or `read_write`, max 8 per session, attached at session create only |
| Memory | `mem_` | one text document at a `path`, ≤100 KB, `409` on path conflict, optimistic concurrency via `content_sha256` |
| Version | `memver_` | **immutable snapshot per mutation**: `operation` ∈ created/modified/deleted, `created_by` actor (session / api / user), filterable by session or API key, full content retrievable |

And the operation we do not have:

> **`redact`** — clears `content`, `content_sha256`, `content_size_bytes` and `path`
> on a historical version while preserving the actor and timestamps.

It exists because of an explicit warning in the docs: *"Never store credentials, API
keys, or tokens in memory stores… a key written once is replayed into every later
session that mounts the store."*

---

## 3. Lode Coding

A single ore-vein of project knowledge, maintained by the assistant rather than by hand.

```
lode/
  overview.md          # mandatory minimum
  <area>/overview.md   # per-area, for larger projects
  <area>/<topic>.md
  tmp/                 # gitignored scraps
```

**Session protocol:** seed (tag the relevant lode files) → state the goal →
orchestrate → review and test → *then* switch to chat mode and ask what changed,
which entries are affected, and what new entry is needed → commit.

**Stance:** the human keeps architectural ownership. *"Only accept what you are
prepared to own."* Zero tooling by design — *"a philosophy and process, not a
technology."*

**Failure mode:** nothing bounds it. No admission test, no size ceiling, no staleness
check, no retrieval discipline — you tag files by hand. Markdown sprawl is not a risk
of Lode; it is the expected steady state.

---

## 4. LeanContext

The same idea with the parts Lode leaves out actually specified.

| Tier | File | Rule |
|---|---|---|
| Constitution | `CONSTITUTION.md` | always loaded, **hard stops only**, 30–50 lines, ceiling 100. Test: *if the agent misses this once, is the damage irreversible?* |
| Manifest | `MANIFEST.md` | routing table, 10–15 lines, **pointers only, never content** |
| Supply | `.context/*.md` | pulled on demand, one topic, <150 lines, **invisible knowledge only** |

**Four rules:** Workbench (only what the operation needs), Shadow Board (perceive
available context without holding it), Pull (load on demand, not at session start),
Andon (stop and surface uncertainty rather than guess).

**Knowledge boundary** — the sharpest thing in either framework:

| Out | In |
|---|---|
| file layout, naming conventions, dependency structure, build commands | why a decision was made, what was tried and rejected, external constraints the code cannot encode |

**Compaction/demotion:** *"The constitution will bloat. This is certain."* Every line
gets re-tested; failures move down a tier. Plus staleness checks, dead-entry audits,
and ADRs in `.context/decisions/` with `Status: Accepted | Superseded | Deprecated`.

**Multi-repo:** a child constitution may not contradict the parent; hard stops flow
downward; a cloned child `curl`s a snapshot of the parent, with the warning that
*"a stale parent constitution is worse than none."*

---

## 5. Head to head

| Axis | Lode | LeanContext | CLAUDE.md + rules | Auto memory | CMA stores | **pi-skills memory** |
|---|---|---|---|---|---|---|
| Substrate | md tree | md, 3 tiers | md hierarchy | md dir | server docs | **SQLite statements** |
| Authorship | agent | human | human | Claude | either | agent |
| Scope | repo | repo + org | 4 layers | git repo | workspace | **graph column** |
| Always loaded | whatever you tag | constitution + manifest | all ancestors | `MEMORY.md` ≤200 lines | mount note | stable policy only |
| Lazy tier | none | `.context/` by manifest | `paths:` globs | topic files | files on mount | per-turn projection |
| Admission test | none | **knowledge boundary** | "would a new teammate need it" | "not derivable from repo" | none | *(to add)* |
| Size discipline | none | 30/15/150 lines | 200-line target | **enforced, errors on write** | 100 KB/doc | `{r} NOTE_BYTES` B/note |
| History | git per file | manual `Status:` | git per file | `modified` stamp | **per-document versions** | **per-field `as_of`** |
| Erasure | `rm` | `rm` | `rm` | `rm` | **`redact`** | *(missing)* |
| Staleness | none | "automate what you can" | none | none | none | *(computable)* |
| Query surface | grep | grep | grep | grep | list + filters | **SQL + FTS + graph** |

The axis that decides cost per session is *always loaded*. Ours is the only design
where that number is a measured constant rather than a target the author is asked
to respect.

```{r fig-budget, fig.height=5}
ours <- (MEM_POLICY + GOAL_POLICY) / 1024
proj <- (WAKE_ROWS * 120 + MATCH_ROWS * NOTE_BYTES) / 1024

d <- data.frame(
  lab = c("Lode Coding",
          "auto memory MEMORY.md",
          "CLAUDE.md (project)",
          "LeanContext tiers",
          "pi-skills stable policy",
          "pi-skills turn projection"),
  kb  = c(NA, 25, 12, 3.9, ours, proj),
  src = c("no bound", "documented cap", "derived", "derived", "measured", "bounded"),
  stringsAsFactors = FALSE
)
d$hi <- grepl("^pi-skills", d$lab)
XM <- 28

blank_plot(c(-13.5, XM + 15), c(0.3, nrow(d) + 0.9), mar = c(3.4, 0.2, 0.2, 0.2))
axis(1, at = seq(0, 25, 5), col = BASE, col.axis = MUT, col.ticks = BASE,
     cex.axis = 0.86, lwd = 1, line = -1.2, tck = -0.02)
text(12.5, -0.55, "KB loaded into every session, before any work", col = MUT, cex = 0.88)
abline(v = seq(5, 25, 5), col = GRID, lwd = 1)

for (i in seq_len(nrow(d))) {
  y <- nrow(d) - i + 1
  text(-0.7, y, d$lab[i], adj = 1, col = if (d$hi[i]) INK else INK2, cex = 0.96)
  if (is.na(d$kb[i])) {
    rect(0, y - 0.26, XM - 2.2, y + 0.26, col = "#f4f3ee", border = BASE, lty = 2)
    arrows(XM - 2.2, y, XM + 1.2, y, length = 0.09, col = MUT, lwd = 1.6)
    text(XM - 3.0, y, "no bound", col = INK2, cex = 0.88, adj = 1)
  } else {
    rect(0, y - 0.26, d$kb[i], y + 0.26,
         col = if (d$hi[i]) BLU else BLU200,
         border = if (d$hi[i]) BLU550 else BLU200, lwd = 1)
    text(d$kb[i] + 0.5, y, sprintf("%.1f KB · %s", d$kb[i], d$src[i]),
         adj = 0, col = INK2, cex = 0.88)
  }
}
segments(0, 0.55, 0, nrow(d) + 0.45, col = BASE)
```

*Derived* rows convert the framework's own line ceiling at 60 bytes/line
(CLAUDE.md 200 lines; LeanContext 50 + 15). *Measured* reads the two policy strings
straight out of `extensions/memory/index.ts` and `extensions/goals/index.ts`.
*Bounded* is `{r} WAKE_ROWS` frontier rows plus `{r} MATCH_ROWS` matches at the
`{r} NOTE_BYTES`-byte note ceiling — the worst case, not an average.

Two further conclusions fall out of the table:

1. On **history and query**, nothing here is close to what the statement log does.
2. On **erasure**, we are the weakest system in the table. Append-only means a
   leaked token is permanent and `as_of` faithfully replays it forever.

---

## 6. What integration means, concretely

### 6.1 Option A — Lode as files (rejected)

Add `lode/lode-map.md`, `summary.md`, `terminology.md`, `plans/`, `tmp/`, plus an
AGENTS.md section per repo. Rejected for three stated reasons: markdown sprawl; no
history beyond per-file git; and text files are not more portable than one SQLite
file with a query surface. Nothing below revisits this.

### 6.2 Option B — Lode as vocabulary inside `memory` (recommended)

**Zero new tables.** A lode entry is a set of statements under a distinct stanza
namespace, so it never enters `note_statement`, the wake frontier, or the summary tree.

```
graph      project:github.com/RGenomicsETL/duckhts
subject    lode:entry/vendoring
```

| Predicate | Value |
|---|---|
| `lode:kind` | `hardstop` · `decision` · `constraint` · `invariant` · `term` · `data` · `gap` |
| `lode:title` | short name |
| `lode:text` | one line, ≤512 bytes |
| `lode:authority` | repo paths that win on conflict |
| `lode:covers` | globs — used for **both** staleness and retrieval |
| `lode:verifiedRev` / `lode:verifiedNote` | git rev + evidence sentence |
| `lode:status` | `active` · `retired` · `gap` |
| `lode:supersedes` / `lode:relatesTo` / `lode:inherits` | object edges, walkable with recursive SQL |

Note the kinds that were **cut**: `area` and `interface`. Both invite exactly the
inferable description LeanContext bans. Every surviving kind is non-inferable by
construction, and `kind=decision` fails the write without a rejected-alternatives
value — the ADR shape enforced rather than templated.

| Concern | Mechanism |
|---|---|
| Read the map before searching | `SELECT` over active entries in the project graph — cannot go stale, never hand-maintained |
| Per-claim history | free from `statement_history` |
| "What did we believe in June" | free from `as_of_statement` |
| Retirement | append `lode:status = retired`; the entry stays inspectable |
| Org-wide rules | `lode:inherits` edge to an org graph; no snapshot, no drift |
| Cross-machine | `memory export --graph project:<id>` → one sorted JSONL at the repo root; `memory import` replays it as one transaction |

### 6.3 Option C — LeanContext as classification (adopt alongside B)

Their three tiers are not three files for us; they are three mechanisms we own.

| LeanContext tier | Their form | Our form |
|---|---|---|
| Constitution | `CONSTITUTION.md`, symlinked to `AGENTS.md` | stable system-prompt policy + `sounkou-engineering-style`; per-repo `kind=hardstop` entries |
| Manifest | hand-maintained table | the lode map, built by `SELECT` |
| Supply | `.context/*.md` | skill progressive disclosure, FTS recall, `lode open` |
| Workbench rule | stated | `context-budget`, enforced in bytes |
| Shadow Board rule | stated | Pi's skill listing |
| Pull rule | manifest discipline | per-turn projection built from the prompt |
| Session discipline | prose | `goals`, with evidence-gated completion |
| Compaction | manual demotion | note stream + summary tree (`nap` / `zoom`) — no equivalent on their side |

Three of their ideas are genuinely new to us:

1. **Knowledge boundary as a write gate.** Refuse entries that restate what the code shows.
2. **Gaps as rows.** Their bootstrap leaves `<!-- FILL IN: why was this chosen? -->` in a file; ours becomes `lode:status = gap` — a question with no answer, surfaced in the projection until answered. The Andon Rule as queryable state.
3. **Demotion, not deletion.** A hardstop failing the irreversibility test gets `lode:kind` re-appended as `constraint`; identity and history survive.

### 6.4 What Claude Code contributes

| Borrowing | Their implementation | Ours |
|---|---|---|
| **Redaction** | `redact` clears content, keeps actor + timestamps | new `redact` op: null the `value` at `(transaction_id, ordinal)`, keep the row, log the erasure as its own transaction with a reason |
| **Glob triggering** | `.claude/rules/` with `paths:` loads on matching file reads | entries whose `lode:covers` match files touched this turn join the projection — the Pull Rule, using a field we already store |
| **Enforced budget** | index over 200 lines / 25 KB → error telling Claude to rewrite it | `entry` write fails when the hardstop cap is exceeded, instead of silently dropping projection rows |
| **Age on every fact** | `modified` frontmatter stamp | render `transaction_time` (or `verified N commits ago`) in the projection — a claim without its date invites over-trust |
| **Placement** | CLAUDE.md is a user message *after* the system prompt, even for managed policy | keep hardstops in the per-turn projection, **not** the cache prefix |

That last row reverses an earlier recommendation of mine, and it comes with a free
property: Claude Code must specially re-read and re-inject project CLAUDE.md after
`/compact`, and admits nested files and `paths:` rules are *not* re-injected — a
projection rebuilt from SQL every user turn simply reappears.

---

## 7. Implementation delta

| Slice | Change | Files |
|---|---|---|
| 1 | `entry()`, `retire()`, `redact()`; `as_of_lode_field` / `as_of_lode_entry` views; `lode_map` / `lode_entry` queries; tool ops `entry` / `retire` / `redact` / `lode`; project-graph resolution from the normalized git remote | `store.ts`, `session.sql`, `operations.sql`, `index.ts` |
| 2 | Projection section (map + task matches + `covers` matches + open gaps + age), FTS union over lode text, status line | `index.ts`, `operations.sql` |
| 3 | Staleness: `verifiedRev`, one `git diff --name-only <rev> HEAD` per distinct rev, glob match in Node; unknown rev → `unknown`, never "fresh"; `verify` requires an evidence note | `index.ts` |
| 4 | `export` / `import` round trip, single sorted JSONL | `store.ts`, `index.ts` |
| 5 | ~5 lines of lode policy in the stable memory prompt; a short "project knowledge" section in `sounkou-engineering-style`. **No new skill.** | `index.ts`, `skills/` |
| 6 | Description entry at `README.Rmd:75`, `npm run render:readme`, CHANGELOG | `README.Rmd`, `CHANGELOG.md` |

Tests to add in `memory.test.ts`: field supersede + `as_of` at an older transaction;
retire hides from the map but keeps history; **note stream and summary tree
unaffected by lode statements**; redact nulls content but preserves the transaction;
export → import is idempotent; FTS finds an entry.

No `SCHEMA_VERSION` bump: no new tables, and the new views carry new names.

---

## 8. Open decisions

1. **Multi-valued `lode:covers`.** Slot semantics means one statement per predicate. A space-separated glob list keeps supersede exact and diffs readable; set membership via object nodes would need tombstones. Proposed: delimited value, capped.
2. **Org graph naming.** `org:RGenomicsETL` versus a normalized org URL. Affects `lode:inherits` and the export.
3. **Where the export lives.** `lode.jsonl` at the repo root versus `.pi/lode.jsonl`.
4. **Redaction scope.** Value only, or also subject/predicate when those leak — a path can be the secret.

---

*Rendered with litedown `{r} as.character(packageVersion("litedown"))`.
Figures are computed from the sources in this repository; re-render with*
`Rscript -e 'litedown::fuse("docs/memory-lode-design.Rmd")'`.
